MSstatsPTM statistical relative quantification of post-translational modifications in global proteomics experiments \textbf{Supplementary Information}

### 2022-07-15 main/MSstatsPTM_Main.tex
replaced ttest with anova

The technological difficulties in PTM identification and quantification increase the uncertainty and the variation in the data, and challenge the downstream statistical analyses. Frequently data from these experiments are analyzed using statistical methods that were not originally designed for this task. Researchers use methods such as $t$-test\cite{Kalpic:2011}, Analysis of Variance\cite{girden:1992}, or Limma\cite{Ritchie_15a}, by taking as input the intensity ratios of modified and unmodified peptide features, and comparing the mean abundance of different PTM sites. Such approaches do not fully account for all the sources of uncertainty. As the result, these approaches are either not directly applicable to experiments with non-trivial designs (such as experiments with multiple conditions, paired and time course designs, and experiments with labeling), or require the analysts to exercise non-trivial statistical expertise.

@@

\todo{OV: even though we say 't-test', the model that we describe is in fact an ANOVA (i.e., nothing limits us to 2 conditions. Should we change it to ANOVA directly?)}

\noindent {\bf Two-sample $t$-test based on modified log$_2$-intensities}

@@

The basic two-sample $t$-test~\cite{kutner_etal_04a} is based on the model
\begin{eqnarray}
y^*_{ijk} = \hat{\mu}^{*}_{i} + \epsilon^{\ast}_{ij}, \ \epsilon^{\ast}_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^{*2}),\ i=1,\ldots,I,\ j=1,\ldots,J,\ k=1,\ldots,K
\label{eq:ttest_model}
\end{eqnarray}
The model allows us to estimate $\hat{\Delta}$ and its standard error.
Based on the model, the approach tests the null hypothesis in \eqref{conv_null_hyp}, by comparing the test statistic against the Student distribution with $df=2J-2$ degrees of freedom in balanced designs.
Unfortunately, this approach is fundamentally flawed as it does not account for the confounding between changes in the PTM abundance and the overall changes in the abundance of the unmodified protein.

@@

\noindent {\bf Two-sample $t$-test based on ratios of modified and unmodified log$_2$-intensities}

\medskip \noindent 
The basic $t$-test can be extended to account for the confounding of changes in PTM abundance and overall changes in protein abundance~\cite{Schwammle2015,THOMAS2020,Mertins:2013}. Typically this is done by first summarizing the feature intensities in each run, separately for modified and unmodified features, e.g. with the sum on the original scale, dividing the summary of the PTM abundance by the summary of the protein abundance and then taking a log. Specifically, for condition $i$ and biological replicate $j$, the approach calculates
\begin{eqnarray}
u_{ij} = \log \left( \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} \right) - \log \left( \sum_{l=1}^{L} 2^{y_{ijl}} \right)
\label{eq:adj_estimation}
\end{eqnarray}
The approach then models these values as
\begin{eqnarray}
u_{ij} = (\mu^{\ast}_{i}-\mu_i) + \epsilon^{\prime}_{ij}, \text{where } \epsilon^{\prime}_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^{\prime 2}), i \in \{1, 2\}
\label{eq:ttest_model}
\end{eqnarray}
The model allows us to estimate $\hat{\Delta}_{adj}$ and its standard error.
Based on this model, we can test the more relevant null hypothesis in~\eqref{null_hyp}, by comparing the test statistic against the Student distribution with \todo{$df=2J-2$} \todo{this should be the same df as the previous section?} degrees of freedom in balanced designs.

Although effective, the approach is somewhat simplistic. It is not applicable to experimental designs with more complex sources of biological and technological variation, such as experiments with repeated measurements, experiments with multiple batches or experiments with multi-run multiplexing. Since \eqref{adj_estimation} performs the adjustment on the replicate level, the experiment must contain a matching number of replicates in both the modified and unmodified runs. Technological artifacts such as missing values further undermine the calculation of $u_{ij}$ in~\eqref{adj_estimation}. Finally, their is no self contained, straightforward implementation of the method, such as in the form of a coding package, requiring a more challenging manual application.

@@

\medskip \noindent The two versions of $t$-test \todo{OV: ANOVA?} above can be expanded with Empirical Bayes moderation in Limma~\cite{ Ritchie_15a, Schwammle2015, Smyth:2004, Smyth:2005, Zhu,Chappell:2021}. In particular, the ratio-based version which models the ratios $\hat{u}_{ij}$ in~\eqref{adj_estimation} can be used to answer the null hypothesis in \eqref{null_hyp}. Limma uses the transformed inputs to fit a linear model with a fixed condition effect for each PTM.

\begin{eqnarray}
& u_{ij} = \mu + Condition_i + \epsilon_{ij}, & \\ 
& \text{where } \sum\limits_{i=1}^{I} Condition_i = 0, \epsilon_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^2) & \nonumber
\label{eq:limma_model}
\end{eqnarray}

@@

The $MSstats$ framework has a number advantages over the methods above. 
First, unlike \todo{t-test (or ANOVA?) and $Limma$}, the family of linear mixed effects models is applicable to many complex experimental designs, including designs with multiple sources of variation, and unbalanced designs. 
Second, the approach is applicable to various data acquisition types, including label-free DDA and DIA, and experiments with TMT labeling, and is compatible with various data processing tools such as Skyline, Spectronaut, MaxQuant, Progenesis, Proteome Discoverer, and OpenMS. Third, the run-level summarization has been shown to be more robust to outliers and missing values as compared to feature averages~\cite{Choi:2014}. 
Finally, the custom $MSstats$ and $MSstatsTMT$ implementation accounts for potential data artifacts, and is numerically scalable and stable, and is available through both command line and a dedicated graphical user interface.

@@

The evaluations were based on the $MSstatsPTM$, and were performed without imputing missing values or using Empirical Bayes moderation. $MSstatsPTM$ was compared with \todo{$t$-test?} and Limma. We did not compare to IsobarPTM because it is only applicable to TMT experiments, meaning it could not be applied to our ground truth data. All the evaluations were done at the adjusted p-value cutoff of $\alpha=.05$. More details are in {\bf Supplementary Sec. \ref{supp-sec:experiments}}.

@@

\figref{sim1_fdr} illustrated the high FDR observed on the simulated datasets when not accounting for the overall changes in protein abundance. After the adjustment, the FDR of all methods improved. In a clean simulation, recall and accuracy of $MSstatsPTM$ and $Limma$ performed similarly. As expected when not using Empiricial Bayes moderation in MSstatsPTM, Limma performed slightly better in simulations with fewer biological replicates. In contrast, the performance of the two-sample $t$-test method lagged behind. This was due to the fact that the $t$-test only uses data within the groups of interest while ignoring the remainder of the data. In contrast, $MSstatsPTM$ and Limma leverage information across comparisons, resulting in improved power in simulations with more comparisons. Additionally, since this simulation is a group comparison design that doesn't require many variance components, it is favorable to Limma and t-test

@@

In simulations with missing values and few features, $MSstatsPTM$ outperformed $t$-test and $Limma$. Changes in unmodified protein level still needed to be accounted for to control the FDR. Once controlling for changes in the unmodified protein, the proposed method outperformed the other methods, as seen in~\figref{sim2_acc}. The proposed method calibrated model accuracy and recall rate well, even when the number of replicates were low. The proposed approach can handle low feature counts due to using Tukey's Median Polish for summarization. The existing methods summation approach requires more observations to correctly summarize the intensities in the run, whereas TMP is robust to few measurements.

Additionally, when comparing the fold change estimation across all modified peptides, the proposed method showed a tighter distribution of estimated fold changes around the true fold change. Specifically, the inner quartile range (IQR) of the estimated fold change for the proposed method was on average 21.8\% smaller than $Limma$ IQR and 10.4\% smaller than $t$-test's IQR. This means that while the mean of the estimated fold changes was generally correct for all methods, the proposed approach correctly estimated the fold change more often across all PTMs. This fold change comparison can be seen in {\bf Supplementary Figure~\ref{supp-fig:fc_boxplot}}.

@@

In this experiment all models incorrectly estimated the fold change of the modified spike-in peptides before adjusting for changes in unmodified protein abundance. After adjustment, the spike-in peptides' fold change was generally in line with expectation in all methods, however the distribution of estimated fold changes was visibly wider (\figref{spikein_boxplot}). Of the three approaches, the proposed method showed the tightest distribution around the true log fold change. Comparing the IQR of the spike-in peptide's log fold change, the proposed method's IQR was 36.78\% smaller than $Limma$ and 32.98\% smaller than $t$-test's. While Limma and $t$-test generally estimated the correct fold change, the proposed method's estimation was more consistent across all modifications.

In~\figref{spikein_prop_volcano} we can clearly see the fold change of the red labeled spike-in peptides was only correctly estimated when accounting for changes in the unmodified protein abundance. Additionally, the background peptides, serving as the null model, show many false positives before adjustment. After adjustment the number of false positives substantially decreased. Specifically, for the proposed method, the number of false positives went from 20.88\% to 1.84\% after adjustment was applied. While the proposed method and $Limma$ both correctly estimated the fold change of the spike-in peptides, using $Limma$ resulted in many large adjusted p-values. Using $Limma$ would have resulted in missing the majority of differential PTMs. This experiment introduced more variance components, which is challenging for the existing methods and therefore the difference in the results is more pronounced than in the simulations.

@@

The results of these experiments are summarized in~\figref{venn_diagrams} and~\figref{profile_plots}. \figref{data4_venn_diagram} and~\figref{data5_venn_diagram} show the number of differentially abundant modified peptides before and after adjustment. Adjusting for changes in the unmodified protein caused more PTMs to change from differentially abundant to non-differentially abundant than vice versa. A question that must be addressed is if the decrease in differentially abundant peptides is due to the increased variance that comes from adjustment. This was tested by looking for modified peptides whose adjusted log fold change was within 10\% of the unadjusted log fold change but became statistically insignificant after adjustment. In other words, the fold change was the same between models but variance increased. When the test was applied to Dataset 4, only one PTM became non-differentially abundant due to an increase in variance. When applied to Dataset 5, 548 PTMs became non-differentially abundant due to an increase in variance (only 3.4\% of all PTMs that became became non-differentially abundant). Thus we can conclude that the drop off in differentially abundant PTMs was due to changes in global protein abundance, not simply increased model variance.

@@

Conversely, cases where the unmodified protein masks changes in the PTM are more infrequent. In this case, the modification and unmodified protein exhibit contradictory changes in abundance, moving in opposite directions. One such case can be seen in~\figref{data4_profile_plot}. Luchetti et al.~\cite{LUCHETTI2021} showed that $GSDMD$ was actively degraded when IpaH7.8 expression was induced by dox treatment. Our reanalysis confirmed this, with the $GSDMD$ protein being down-regulated when Dox treatments reached the 4 and 6 hour marks. Conversely, ubiquitination of $GSDMD$ at site $K62$ up-regulated abundance between the same conditions. This up-regulation was originally confounded by the down-regulation of unmodified $GSDMD$, and made the modification appear to have little change between no Dox and Dox 4 and 6 hour conditions. The proposed approach accounted for this confounding and the modification was shown to be differentially abundant, with an absolute fold change of $1.415$ between the Dox 1 hour and Dox 4 hour conditions. This was an interesting result, as the modification contradicts what previous research showed for the unmodified protein. This effect can be thought of as a false negative and would have been challenging to observe without the proposed approach. There are many non-differentially abundant PTMs, and having to comb through each one individually to determine false negatives is generally an unfeasible task. The proposed approach removes the confounding on an experimental level, revealing interesting PTMs that are entirely masked due to changes in the unmodified protein.

@@

The number of differentially abundant PTMs before and after adjustment are shown in~\figref{data6_plots}. After adjusting for changes in the unmodified protein, there were fewer differentially modified peptides. However, this was mainly due to not having a global profiling run. When looking only at PTMs with a corresponding unmodified protein, the number of differentially modified PTMs exhibited a small drop, but was more consistent. As above, we tested if this drop in differentially abundant PTMs was due to an increase in variance. Here only 25 PTMs became insignificant due to increased variance. When there was no global profiling run the methods were less effective at removing confounding with the unmodified protein. This was entirely due to not being able to perform adjustment without a corresponding global protein. In this case adjustment caused us to toss out a large number of the PTMs in the original model. In cases without a global profiling run, the proposed method can still be used, but it should be used in conjunction with the unadjusted model so that information is not lost.

@@

We proposed a general statistical modeling framework and implementation for PTM characterization. The framework is designed for bottom-up MS workflows, which are characterized with variations from multiple confounded sources, frequent missing data, and associated uncertainty in the conclusions. The framework is general and is applicable to a variety of experimental designs. It outperforms the ad-hoc methods underlying $t$-test and Limma, and yields accurate results in the broad type of experimental circumstances, including the presence of missing values, changes in protein abundance, few representative peptides, and different acquisition methods. The framework allows us to plan for subsequent experiments, and choose the appropriate number of replicates in consideration of adjustment with respect to protein abundance. The implementation allows for straightforward application of the methods discussed and allows for reproducible experimental analysis.

@@

\begin{figure}[ht]
\centering
\begin{subfigure}[c]{0.825\linewidth}
\includegraphics[width=1\textwidth]{images/sim1_FDR_all_models.png}
\caption{}
\label{fig:sim1_fdr}
\end{subfigure}
\begin{subfigure}[c]{0.825\linewidth}
\includegraphics[width=1\textwidth]{images/sim3_Accuracy.png}
\caption{}
\label{fig:sim2_acc}
\end{subfigure}
\caption{Dataset 1 \& 2: Computer simulation. a) All the considered methods in the first computer simulation correctly calibrated FDR when adjusting for changes in protein abundance. In comparison, the methods without accounting for the protein-level changes resulted in off-target, high false positive rates. b) The advantage of using the proposed approach was apparent when including limited observations and missing values. Looking at accuracy, the proposed method outperformed Limma and $t$-test in nearly every model.
}
\label{fig:computer_sim}
\end{figure}

\begin{figure}[ht]
\centering
\begin{subfigure}[c]{0.825\linewidth}
\includegraphics[width=1\textwidth]{images/spike_in_fc.png}
\caption{}
\label{fig:spikein_boxplot}
\end{subfigure}
\begin{subfigure}[c]{0.825\linewidth}
\includegraphics[width=1\textwidth]{images/spike_in_volcano.png}
\caption{}
\label{fig:spikein_prop_volcano}
\end{subfigure}
\caption{Dataset 3: Spike-in benchmark - Ubiquitination - Label-free. a) Before adjustment the fold change of the spike-in peptides' were systematically different from the expected fold change in all models. After adjustment, this systemic difference was removed, however the inner quartile range of the Limma and $t$-test models was wider than the proposed method. b) Before adjustment the spike-in peptides (colored red) did not follow the expected log fold change; after adjustment, the spike-in peptides were more in line with expectation. Using Limma, the spike-in peptides followed the expected log fold change after adjustment, however the majority of spike-in peptides did not have a significant adjusted p-value.}
\label{fig:spikein_volcano}
\end{figure}

>>>>

The technological difficulties in PTM identification and quantification increase the uncertainty and the variation in the data, and challenge the downstream statistical analyses. Frequently data from these experiments are analyzed using statistical methods that were not originally designed for this task. Researchers use methods such as $t$-test\cite{Kalpic:2011}, Analysis of Variance (Anova)\cite{girden:1992}, or Limma\cite{Ritchie_15a}, by taking as input the intensity ratios of modified and unmodified peptide features, and comparing the mean abundance of different PTM sites. Such approaches do not fully account for all the sources of uncertainty. As the result, these approaches are either not directly applicable to experiments with non-trivial designs (such as experiments with multiple conditions, paired and time course designs, and experiments with labeling), or require the analysts to exercise non-trivial statistical expertise.

@@

\noindent {\bf Anova based on modified log$_2$-intensities}

@@

The basic Anova~\cite{kutner_etal_04a} is based on the model
\begin{eqnarray}
y^*_{ijk} = \hat{\mu}^{*}_{i} + \epsilon^{\ast}_{ij}, \ \epsilon^{\ast}_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^{*2}),\ i=1,\ldots,I,\ j=1,\ldots,J,\ k=1,\ldots,K
\label{eq:ttest_model}
\end{eqnarray}
The model allows us to estimate $\hat{\Delta}$ and its standard error.
Based on the model, the approach tests the null hypothesis in \eqref{conv_null_hyp}, by comparing the test statistic against the Student distribution with $df=IJ-I$ degrees of freedom in balanced designs.
Unfortunately, this approach is fundamentally flawed as it does not account for the confounding between changes in the PTM abundance and the overall changes in the abundance of the unmodified protein.

@@

\noindent {\bf Anova based on ratios of modified and unmodified log$_2$-intensities}

\medskip \noindent 
The basic Anova can be extended to account for the confounding of changes in PTM abundance and overall changes in protein abundance~\cite{Schwammle2015,THOMAS2020,Mertins:2013}. Typically this is done by first summarizing the feature intensities in each run, separately for modified and unmodified features, e.g. with the sum on the original scale, dividing the summary of the PTM abundance by the summary of the protein abundance and then taking a log. Specifically, for condition $i$ and biological replicate $j$, the approach calculates
\begin{eqnarray}
u_{ij} = \log \left( \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} \right) - \log \left( \sum_{l=1}^{L} 2^{y_{ijl}} \right)
\label{eq:adj_estimation}
\end{eqnarray}
The approach then models these values as
\begin{eqnarray}
u_{ij} = (\mu^{\ast}_{i}-\mu_i) + \epsilon^{\prime}_{ij}, \text{where } \epsilon^{\prime}_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^{\prime 2}),\ i=1,\ldots,I,\ j=1,\ldots,J
\label{eq:ttest_model}
\end{eqnarray}
The model allows us to estimate $\hat{\Delta}_{adj}$ and its standard error.
Based on this model, we can test the more relevant null hypothesis in~\eqref{null_hyp}, by comparing the test statistic against the Student distribution with $df=IJ-I$ degrees of freedom in balanced designs.

Although effective, the approach is somewhat simplistic. It is not applicable to experimental designs with more complex sources of biological and technological variation, such as experiments with repeated measurements, experiments with multiple batches or experiments with multi-run multiplexing. Since \eqref{adj_estimation} performs the adjustment on the replicate level, the experiment must contain a matching number of replicates in both the modified and unmodified runs. Technological artifacts such as missing values further undermine the calculation of $u_{ij}$ in~\eqref{adj_estimation}. Finally, there is no self contained, straightforward implementation of the method, such as in the form of a coding package, requiring a manual application.

@@

\medskip \noindent The two versions of Anova above can be expanded with Empirical Bayes moderation in Limma~\cite{ Ritchie_15a, Schwammle2015, Smyth:2004, Smyth:2005, Zhu,Chappell:2021}. In particular, the ratio-based version which models the ratios $\hat{u}_{ij}$ in~\eqref{adj_estimation} can be used to answer the null hypothesis in \eqref{null_hyp}. Limma uses the transformed inputs to fit a linear model with a fixed condition effect for each PTM.

\begin{eqnarray}
& u_{ij} = \mu + Condition_i + \epsilon_{ij}, \text{where } \sum\limits_{i=1}^{I} Condition_i = 0, \epsilon_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^2)
\label{eq:limma_model}
\end{eqnarray}

@@

The $MSstats$ framework has a number advantages over the methods above. 
First, unlike Anova and $Limma$, the family of linear mixed effects models is applicable to many complex experimental designs, including designs with multiple sources of variation, and unbalanced designs. 
Second, the approach is applicable to various data acquisition types, including label-free DDA and DIA, and experiments with TMT labeling, and is compatible with various data processing tools such as Skyline, Spectronaut, MaxQuant, Progenesis, Proteome Discoverer, and OpenMS. Third, the run-level summarization has been shown to be more robust to outliers and missing values as compared to feature averages~\cite{Choi:2014}. 
Finally, the custom $MSstats$ and $MSstatsTMT$ implementation accounts for potential data artifacts, and is numerically scalable and stable, and is available through both command line and a dedicated graphical user interface.

@@

The evaluations were based on the $MSstatsPTM$, and were performed without imputing missing values or using Empirical Bayes moderation. $MSstatsPTM$ was compared with Anova and Limma. We did not compare to IsobarPTM because it is only applicable to TMT experiments, meaning it could not be applied to our ground truth data. All the evaluations were done at the adjusted p-value cutoff of $\alpha=.05$. More details are in {\bf Supplementary Sec. \ref{supp-sec:experiments}}.

@@

\figref{sim1_fdr} illustrated the high FDR observed on the simulated datasets when not accounting for the overall changes in protein abundance. After the adjustment, the FDR of all methods improved. In a clean simulation, recall and accuracy of $MSstatsPTM$ and Limma performed similarly. As expected when not using Empiricial Bayes moderation in MSstatsPTM, Limma performed slightly better in simulations with fewer biological replicates. In contrast, the performance of Anova lagged behind, especially in cases with higher simulated variance. This was due to Anova's more simplistic variance estimation. In contrast, $MSstatsPTM$ and Limma implement methods to provide better estimates of the variance, resulting in improved power. Finally, since this simulation is a group comparison design that doesn't require many variance components, it is favorable to Limma and Anova. As discussed previously, these methods are not directly applicable to experiments with additional variance components.

@@

In simulations with missing values and few features, $MSstatsPTM$ outperformed Anova and Limma. Changes in unmodified protein level still needed to be accounted for to control the FDR. Once controlling for changes in the unmodified protein, the proposed method outperformed the other methods, as seen in~\figref{sim2_acc}. The proposed method calibrated model accuracy  well, even when the number of replicates were low. $MSstatsPTM$ can handle low feature counts better than the existing methods due to using Tukey's Median Polish for summarization. The existing methods summation approach requires more observations to correctly summarize the intensities in the run, whereas TMP is robust to few measurements.

Additionally, when comparing the fold change estimation across all modified peptides, the proposed method showed a tighter distribution of estimated fold changes around the true fold change. Specifically, the inner quartile range (IQR) of the estimated fold change for the proposed method was on average 32.5\% smaller than Limma and Anova's IQR. This means that while the mean of the estimated fold changes was generally correct for all methods, the proposed approach correctly estimated the fold change more often across all PTMs. This fold change comparison can be seen in {\bf Supplementary Figure~\ref{supp-fig:fc_boxplot}}.

@@

In this experiment all models incorrectly estimated the fold change of the modified spike-in peptides before adjusting for changes in unmodified protein abundance. After adjustment, the spike-in peptides' fold change was generally in line with expectation in all methods, however the existing method's distribution of estimated fold changes was visibly wider (\figref{spikein_boxplot}). Of the three approaches, the proposed method showed the tightest distribution around the true log fold change. Comparing the IQR of the spike-in peptide's log fold change, the proposed method's IQR was 32.86\% smaller than Limma and Anova's IQR. While Limma and Anova overall estimated the correct fold change, they were inconsistent in their estimation, resulting in a higher error rate. The proposed method's estimation was more consistent, with a smaller error. As discussed in the previous section, the summarization methods of the proposed methods excel when the number of observations are low. This was the case here, with each PTM only averaging $1.37$ features. The existing approaches used a non-robust summarization method, that was very sensitive to few observations, causing their fold change estimation to be more inconsistent, with a larger error compared to the proposed method.

In~\figref{spikein_prop_volcano} we can clearly see the fold change of the red labeled spike-in peptides was only correctly estimated when accounting for changes in the unmodified protein abundance. Additionally, the background peptides, serving as the null model, show many false positives before adjustment. After adjustment the number of false positives substantially decreased. Specifically, for the proposed method, the number of false positives went from 20.88\% to 1.84\% after adjustment was applied. While the proposed method and Limma both correctly estimated the fold change of the spike-in peptides, using Limma resulted in many large adjusted p-values. This was due to Limma estimating a 35.7\% larger variance compared to the proposed method when estimating the spike-in peptides. This experiment introduced more variance components, which is challenging for the existing methods and therefore the difference in the results is more pronounced than in the simulations.

@@

The results of these experiments are summarized in~\figref{venn_diagrams} and~\figref{profile_plots}. \figref{data4_venn_diagram} and~\figref{data5_venn_diagram} show the number of differentially abundant modified peptides before and after adjustment. Adjusting for changes in the unmodified protein caused more PTMs to change from differentially abundant to non-differentially abundant than vice versa. A question that must be addressed is if the decrease in differentially abundant peptides is due to the increased variance that comes from adjustment. This was tested by looking for modified peptides whose adjusted log fold change was within 10\% of the unadjusted log fold change but became statistically insignificant after adjustment. In other words, the fold change was the same between models but variance increased. When the test was applied to Dataset 4, only one PTM became non-differentially abundant due to an increase in variance. When applied to Dataset 5, 548 PTMs became non-differentially abundant due to an increase in variance (only 3.4\% of all PTMs that became non-differentially abundant). Thus we can conclude that the drop off in differentially abundant PTMs was due to changes in global protein abundance, not simply increased model variance.

@@

Conversely, cases where the unmodified protein masks changes in the PTM are more infrequent. In this case, the modification and unmodified protein exhibit contradictory changes in abundance, moving in opposite directions. One such case can be seen in~\figref{data4_profile_plot}. Luchetti et al.~\cite{LUCHETTI2021} showed that $GSDMD$ was actively degraded when IpaH7.8 expression was induced by dox treatment. Our reanalysis confirmed this, with the $GSDMD$ protein being down-regulated when Dox treatments reached the 4 and 6 hour marks. Conversely, ubiquitination of $GSDMD$ at site $K62$ up-regulated abundance between the same conditions. This up-regulation was originally confounded by the down-regulation of unmodified $GSDMD$, and made the modification appear to have little change between no Dox and Dox 4 and 6 hour conditions. The proposed approach accounted for this confounding and the modification was shown to be differentially abundant, with an absolute fold change of $1.415$ between the Dox 1 hour and Dox 4 hour conditions. This was an interesting result, as the modification contradicts what previous research showed for the unmodified protein. This effect can be thought of as a false negative and would have been challenging to observe without the proposed approach. There are many non-differentially abundant PTMs, and having to comb through each one individually to determine false negatives is generally an unfeasible task. The proposed approach removes the confounding on an experimental level, revealing modifications of interest that are entirely masked due to changes in the unmodified protein.

@@

The number of differentially abundant PTMs before and after adjustment are shown in~\figref{data6_plots}. After adjusting for changes in the unmodified protein, there were fewer differentially modified peptides. However, this was mainly due to not having a global profiling run. When looking only at PTMs with a corresponding unmodified protein, the number of differentially modified PTMs exhibited a small drop, but was more consistent. As above, we tested if this drop in differentially abundant PTMs was due to an increase in variance. Here only 25 PTMs became insignificant due to increased variance. Without a global profiling run, the method was less effective at removing confounding with the unmodified protein. This was due to not being able to perform adjustment without a corresponding global protein. In this case adjustment caused us to toss out a large number of the PTMs in the original model. In cases without a global profiling run, the proposed method can still be used, but it should be used in conjunction with the unadjusted model so that information is not lost.

@@

We proposed a general statistical modeling framework and implementation for PTM characterization. The framework is designed for bottom-up MS workflows, which are characterized with variations from multiple confounded sources, frequent missing data, and associated uncertainty in the conclusions. The framework is general and is applicable to a variety of experimental designs. It outperforms the ad-hoc methods underlying Anova and Limma, and yields accurate results in the broad type of experimental circumstances, including the presence of missing values, changes in protein abundance, few representative peptides, and different acquisition methods. The framework allows us to plan for subsequent experiments, and choose the appropriate number of replicates in consideration of adjustment with respect to protein abundance. The implementation allows for straightforward application of the methods discussed and allows for reproducible experimental analysis.

@@

\begin{figure}[ht]
\centering
\begin{subfigure}[c]{0.825\linewidth}
\includegraphics[width=1\textwidth]{images/sim1_FDR_all_models.png}
\caption{}
\label{fig:sim1_fdr}
\end{subfigure}
\begin{subfigure}[c]{0.825\linewidth}
\includegraphics[width=1\textwidth]{images/sim3_Accuracy.png}
\caption{}
\label{fig:sim2_acc}
\end{subfigure}
\caption{Dataset 1 \& 2: Computer simulation. a) All the considered methods in the first computer simulation correctly calibrated FDR when adjusting for changes in protein abundance. In comparison, the methods without accounting for the protein-level changes resulted in off-target, high false positive rates. b) The advantage of using the proposed approach was apparent when including limited observations and missing values. Looking at accuracy, the proposed method outperformed Anova and Limma in nearly every model.
}
\label{fig:computer_sim}
\end{figure}

\begin{figure}[ht]
\centering
\begin{subfigure}[c]{0.825\linewidth}
\includegraphics[width=1\textwidth]{images/spike_in_fc.png}
\caption{}
\label{fig:spikein_boxplot}
\end{subfigure}
\begin{subfigure}[c]{0.825\linewidth}
\includegraphics[width=1\textwidth]{images/spike_in_volcano.png}
\caption{}
\label{fig:spikein_prop_volcano}
\end{subfigure}
\caption{Dataset 3: Spike-in benchmark - Ubiquitination - Label-free. a) Before adjustment the fold change of the spike-in peptides' were systematically different from the expected fold change in all models. After adjustment, this systemic difference was removed, however the inner quartile range of the Limma and Anova models was wider than the proposed method. b) Before adjustment the spike-in peptides (colored red) did not follow the expected log fold change; after adjustment, the spike-in peptides were more in line with expectation. Using Limma, the spike-in peptides followed the expected log fold change after adjustment, however the majority of spike-in peptides did not have a significant adjusted p-value.}
\label{fig:spikein_volcano}
\end{figure}

### 2022-07-15 supplementary/ptm_sm.tex
replaced ttest with anova

The results are summarized from Figure \ref{fig:sim1_fpr} to Figure \ref{fig:sim1_recall}. These results include the false discovery rate with and without adjusting for changes in unmodified protein abundance, false positive rate, and overall accuracy. Additionally, the results are compared to simulation 2 described in the next section.

@@

\begin{figure}[h!]
\centering
 \begin{subfigure}{\textwidth}
  \centering
	\includegraphics[width=.8\textwidth]{sim_new/sim1_FDR_all_models}\\
	\caption{All the considered methods in simulation 1 correctly calibrated FDR  when adjusting for changes in protein abundance. In comparison, the methods without accounting for the protein-level changes resulted in off-target, high false discovery rates. The performance of the models without adjustment was much lower than those with adjustment, thus only models with adjustment are compared going forward.}
 \end{subfigure}
 \begin{subfigure}{\textwidth}
  \centering
	\includegraphics[width=.8\textwidth]{sim_new/sim1_FDR}
	\caption{The considered methods with protein adjustment were compared in detail. All three methods with adjustment generally performed similarly in terms of FDR.}
 \end{subfigure}
 \caption{FDR of Simulation 1.}
\label{fig:sim1_fpr}
\end{figure}

@@

\begin{figure}[h!]
\centering
\includegraphics[width=.85\textwidth]{sim_new/spike_in_limma_volcano}
\caption{When modeling the experiment with the Limma method, the spike-in peptides again follow the expected log fold change better after adjusting for changes in protein level. However, while the fold change was more accurate, the majority of spike-in peptides did not have a significant adjusted pvalue. In this case, the known differential peptides were missed by the model. In terms of false positives, the results were very similar to the proposed method, with many false positives before adjustment and fewer after. \label{fig:spike_volcano_limma}}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[width=.85\textwidth]{sim_new/spike_in_ttest_volcano}
\caption{Using the two sample $t$-test, none of the comparisons either before or after adjustment show any significant peptides. With that being said, the fold change of the spike-in peptides was much closer to expectation after adjusting for global protein abundance. \label{fig:spike_volcano_ttest}}
\end{figure}

>>>>

The results are summarized in Figure \ref{fig:sim1_recall}.

@@

@@

\begin{figure}[h!]
\centering
\includegraphics[width=1\textwidth]{sim_new/spike_in_limma_volcano}
\caption{When modeling the experiment with the Limma method, the spike-in peptides again follow the expected log fold change better after adjusting for changes in protein level. However, while the fold change was more accurate, the majority of spike-in peptides did not have a significant adjusted pvalue. In this case, the known differential peptides were missed by the model. In terms of false positives, the results were very similar to the proposed method, with many false positives before adjustment and fewer after. \label{fig:spike_volcano_limma}}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[width=1\textwidth]{sim_new/spike_in_anova_volcano}
\caption{Using the two sample $t$-test, none of the comparisons either before or after adjustment show any significant peptides. With that being said, the fold change of the spike-in peptides was much closer to expectation after adjusting for global protein abundance. \label{fig:spike_volcano_ttest}}
\end{figure}
